---
title: "Team 7 - EPL Football Results (Tidyverse)"
format: html
author: Firdauz B K, Jit Lin, Jonathan, Xiang Hong
editor: visual
toc: true
toc-depth: 2
embed-resources: true
references:
  - title: "Car Specification Dataset 1945-2020"
    url: https://www.kaggle.com/datasets/jahaidulislam/car-specification-dataset-1945-2020
    author: "Jahaidul Islam"
    year: 2023
    description: "This dataset provides detailed specifications of car models ranging from 1945 to 2020, including engine capacity and price, suitable for analytical purposes in automotive industry studies."
---

## Load Libraries

```{r}
library(gt)
library(tidyverse)
library(readr)
```

## Import Dataset

```{r}
team <- read_csv("dataset/epl_teams.csv")
results <- read_csv("dataset/epl_results.csv")
```

# Task 1: Converting EPL Football Results into a League Table

## 1.1 Generate a League Table from the EPL Results

...

## 1.2 Improve Visualization Using `gt` Functions

...

# Task 2: Finding and Visualizing a Real-World Dataset

## Import the dataset

The dataset contains a list of cars manufactured from 1945 to 2020. It includes details such as make, model, year, engine size, fuel type, transmission type, drivetrain, body style, number of doors, etc. The data is from [Kaggle](https://www.kaggle.com/datasets/jahaidulislam/car-specification-dataset-1945-2020) by Jahaidul Islam.
```{r}
cdf <- read_csv("dataset/car_data.csv")
tail(cdf)
nrow(cdf)
```
...

# Task 3: Visualizing Michelson’s Speed-of-Light Measurements

```{r}
data(morley)
morley <- as_tibble(morley)
morley
```

## 3.1 Convert 'Expt' Column to Factoe

```{r}
morley <- morley %>%
  mutate(Expt = as.factor(Expt))
```

## 3.2 Plot 1 – Create a Scatter Plot displaying Experiment Number vs. Measured Speeds

```{r}
morley %>%
  ggplot(aes(x = Expt, y = Speed)) +
  geom_point() +
  labs(title = "Measured Speeds vs Experiment Number",
       x = "Experiment Number",
       y = "Measured Speeds")
```

## Plot 2 – Create a Box Plot Measured Speeds with One Box Experiment

```{r}
morley %>%
  ggplot(aes(x = Expt, y = Speed)) +
  geom_boxplot() +
  labs(title = "Box Plot of Measured Speeds with One Box Experiment",
       x = "Experiment Number",
       y = "Measured Speeds")
```

## Plot 3 – Add a Violin Plot in Plot 2

```{r}
morley %>%
  ggplot(aes(x = Expt, y = Speed)) +
  geom_boxplot() +
  geom_violin() +
  labs(title = "Box Plot of Measured Speeds with One Box Experiment",
       x = "Experiment Number",
       y = "Measured Speeds")
```

## Plot 4 – Create a Faceted Plot Containing 5 Histograms

```{r}
morley %>%
  ggplot(aes(x = Speed)) +
  geom_histogram() +
  facet_wrap(~Expt) +
  labs(title = "Faceted Plot of Measured Speeds",
       x = "Measured Speeds",
       y = "Frequency")
```

## Discuss Your Choice Among the 4 Visualization Plots
